\texorpdfstring{Identifying Software Performance Changes Across\\ Variants and Versions}{Identifying Software Performance Changes Across Variants and Versions}

### 2020-08-29 paper.tex
paper

%%
%% This is file `sample-sigconf.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `sigconf')
%%
%% IMPORTANT NOTICE:
%%
%% For the copyright see the source file.
%%
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigconf.tex.
%%
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%%
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigconf]{acmart}
\usepackage{microtype}
\usepackage[utf8]{inputenc}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{enumerate}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{mathtools} 
\usepackage{subcaption}
\usepackage{caption}
\usepackage{tikz}
\usepackage{mdframed}
\usepackage{wrapfig}
\usepackage{caption}

@@

	\author{Stefan Mühlbauer}
	\affiliation{%
		\institution{Leipzig University,}
		%\city{Doimukh}
		\country{Germany}}

	\author{Sven Apel}
	\affiliation{%
		\institution{Saarland University,}
		%\city{Doimukh}
		\country{Germany}}

	\author{Norbert Siegmund}
	\affiliation{%
		\institution{Leipzig University,}
		%\city{Doimukh}
		\country{Germany}}
	%%
	%%
	%% By default, the full list of authors will be used in the page
	%% headers. Often, this list is too long, and will overlap
	%% other information printed in the page headers. This command allows
	%% the author to define a more concise list
	%% of authors' names for this purpose.
	%\renewcommand{\shortauthors}{Trovato and Tobin, et al.}

@@

	\begin{CCSXML}
		<ccs2012>
		<concept>
		<concept_id>10011007.10010940.10011003.10011002</concept_id>
		<concept_desc>Software and its engineering~Software performance</concept_desc>
		<concept_significance>500</concept_significance>
		</concept>
		<concept>
		<concept_id>10010147.10010257.10010282.10011304</concept_id>
		<concept_desc>Computing methodologies~Active learning settings</concept_desc>
		<concept_significance>500</concept_significance>
		</concept>
		<concept>
		<concept_id>10011007.10011074.10011111.10011113</concept_id>
		<concept_desc>Software and its engineering~Software evolution</concept_desc>
		<concept_significance>500</concept_significance>
		</concept>
		</ccs2012>
	\end{CCSXML}

	\ccsdesc[500]{Software and its engineering~Software performance}
	\ccsdesc[500]{Computing methodologies~Active learning settings}
	\ccsdesc[500]{Software and its engineering~Software evolution}

@@

	%%
	%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
	%% Please copy and paste the code instead of the example below.h
	%%
	%% Keywords. The author(s) should pick words that accurately describe
	%% the work being presented. Separate the keywords with commas.
	\keywords{software performance, software evolution, configurable software systems, machine learning, active learning}

@@

	\subsection{Performance-Influence Models}\label{sec:pimodels}
	Consider the following running example of a database management system (DBMS) with two selectable features: \textsc{Encryption} and \textsc{compression}.
	Either feature adds execution time to the overall performance, but, if both features are selected, the execution time is smaller than the sum of the individual features' contribution to performance: Less data is encrypted if it is compressed beforehand.
	An \textit{interaction} in this setting is the combined effect of features \textsc{Encryption} and \textsc{compression}.
	We can assign each feature and interaction an \textit{influence} that it contributes to the overall performance, if selected. In our example, the individual influences of the two features are positive (increasing execution time), whilst the interaction's influence is negative (decreasing execution time).

@@

	We encode the threshold in a step function $\theta_\tau$:
	\begin{equation}\label{eq:theta}
	\begin{gathered}
	\theta_\tau(a, b) = \begin{cases}
	0,\quad \vert\pi(a) - \pi(b)\vert < \tau\\
	1,\quad \vert\pi(a) - \pi(b)\vert \geq \tau
	\end{cases}\quad a, b \in V, \tau \in \mathbb{R}
	\end{gathered}
	\end{equation}
	The function evaluates to $1$ if the difference between performance $\pi_a$ and $\pi_b$ of commits $a$ and $b$ exceeds the threshold $\tau$ and $0$ if not.

@@

	The threshold $t_\text{CPL}$ is the average change point likelihood over all commits plus a factor $N_\text{CPL}$ times its corresponding standard deviation over all commits. 
	The factor $N_\text{CPL}$ (default value: 3) allows us to filter peaks that do not stand out enough and might be false positives. The greater $N_\text{CPL}$ is, the stricter the filtering of peaks. That is, for each configuration in our learning set, we obtain a set of commits (configuration-specific candidate commits) that represent possible change points.
	To reduce variation among the obtained commits, we cluster the commits using kernel density estimation (KDE). The purpose of KDE in our setting is to estimate the probability mass function of whether a commit is a change point. The local maxima of this KDE, subsequently, represent our set of candidate commits. Note that, if two distinct change points are almost coincident commit-wise, they can be mistakenly identified by our approach as one change point that, subsequently, can be associated with configuration options belonging to the two individual change points.

@@

	For each model (for a candidate commit), we consider a configuration option as associated with a commit, if its influence $c_i$ (i.e., the absolute value of its coefficient) is greater than a threshold:% $t_\text{influence}$:
	\begin{equation}
	\begin{gathered}
	t_\text{influence} = \frac{1}{n_\text{options}} + N_\text{influence} \cdot \underbrace{\sqrt{\frac{1}{n_\text{options}} \cdot \sum \Big(\frac{1}{n_\text{options}} - c_i\Big)^2}}_{\text{standard deviation of } c_i}
	\label{eq:t_influence}
	\end{gathered}
	\end{equation}
	We consider options as influential, if their estimated influence is greater than if each configuration option were equally influential. In addition, we add to this threshold the standard deviation of the influences of all configuration options. In theory, our regression model eliminates non-influential configuration options. If more than configuration option is influential (i.e., two or more options interact) their respective estimated influence should be roughly equal. To avoid false positives in he estimation of influential configuration options, we use the standard deviation and the parameter $N_\text{influence}$ (default value: 1) to filter only estimations with low variation among the estimated influences. If the intercept of a model (i.e., the coefficient not related to any regression model variable) exceeds this threshold, we consider this candidate commit a change point that is not configuration-specific (i.e., it affects all configurations). The outcome is a list of \emph{associations}: pairs of candidate commits and (likely) influential configuration options. 
	% unklar, warum N_influence hier eingeführt wird

@@

	With this split experiment setup, we aim at answering the following two research questions:
	\begin{enumerate}[\textbf{$\text{RQ}_{1}$}:]
		\item Can we \emph{accurately} and \emph{efficiently} identify configuration-specific performance change points? 
		\item Can we \emph{practically} identify configuration-specific change points \emph{in a real-world setting}? 
	\end{enumerate}

	\subsection{Controlled Experiment Setup}
	We break down the research question $\text{RQ}_1$ into three objectives  to study the influence of the size of configurable software systems, change point properties, and measurement effort.

@@

	\begin{table}
		\centering
		\captionof{table}{Parameter ranges for the synthetic experiment.}
		\label{tab:synthetic_experiment_ranges}
		\begin{tabular}{llr}
			\toprule
			\textbf{}                & \textbf{Parameter}        & \multicolumn{1}{r}{\textbf{Range}} \\
			\midrule
			\parbox[t]{20mm}{\multirow{4}{*}{\rotatebox[origin=c]{0}{\parbox{2cm}{\centering\footnotesize Synthesized Systems}}}} & $n_\text{options}$                & 8, 16, 32, 64                \\
			& $n_\text{commits}$                & 1000, 2500                \\
			& $n_\text{changepoints}$           & 1, 2, 5, 10                        \\
			& $p_\text{interaction}$ & 0.5, 0.7, 0.9                     \\
			\midrule
			\parbox[t]{20mm}{\multirow{2}{*}{\rotatebox[origin=c]{0}{\parbox{2cm}{\centering\footnotesize Initialization}}}}   & $N_\text{initial}$       & 2, 5, 10                           \\
			& $n_\text{measurements}$           & 100, 200, 500                      \\
			\bottomrule            
		\end{tabular}
	\end{table}

@@

	\begin{table}
		\centering
		\captionof{table}{Project characteristics for our three subject systems}
		\label{tab:properties}
		\begin{tabular}{lrr}
			\toprule
			\textbf{Name} & \textbf{\#Options} & \textbf{\#Commits}\\
			\midrule
			\textsc{xz} & 16 & 1193\\
			\textsc{lrzip} & 9 & 743 \\
			\textsc{oggenc} & 12 & 935\\
			\bottomrule
		\end{tabular}
	\end{table}

@@

	\subsection{Results}	
	\subsubsection{$\text{RQ}_1$ (Controlled Experiment).}
	We illustrate the results of our first set of experiments in Figures~\ref{fig:results_rq1_accuracy_terminated}, \ref{fig:results_rq1_frequency}, and \ref{fig:results_rq1_degree}. The vast majority of experiments terminated within the limit of 30 iterations, as shown in Figure~\ref{fig:results_rq1_frequency}. A small portion of experiments, however, did not meet our termination criteria (more on that below). From all experiments that terminated in Figure~\ref{fig:results_rq1_accuracy_terminated}, we depict the F1 score after they have terminated, that is, after their last iteration\footnote{Due to space limitations we report precision and recall on the paper's companion Web site.}. For most iterations, the mean F1 score falls around or over 0.7, with the first quartile  only slightly being as low as 0.5 (at iteration 29).  In the grand scheme of things, the vast majority of experiments terminated with reasonably high F1 score. 
	\begin{figure}
		\begin{subfigure}{\linewidth}
			\centering
			\includegraphics[width=\linewidth]{images/accuracy_after_termination.pdf}
			\vspace{-5mm}
			\caption{F1 score of experiments terminating at different iterations.}	
			\label{fig:results_rq1_accuracy_terminated}
		\end{subfigure}
		\begin{subfigure}{\linewidth}
			\centering
			\vspace{1mm}
			\includegraphics[width=\linewidth]{images/iterations_for_termination.pdf}
			\vspace{-5mm}
			\caption{Frequency of experiments terminating at different iterations.}	
			\label{fig:results_rq1_frequency}
		\end{subfigure}
		\begin{subfigure}{\linewidth}
			\vspace{4mm}
			\centering
			\includegraphics[width=\linewidth]{images/max_degree_after_termination}
			\caption{Influence of the number of change points and interaction degree on accuracy}	
			\label{fig:results_rq1_degree}
		\end{subfigure}
		\caption{Result for the synthetic experiment}
		\label{fig:results_rq1:results_rq1_degree}
	\end{figure} 	
	Regarding measurement effort, for small systems with 8 configuration options, we required up to 50\,\% of all possible measurements. For greater systems, the required measurement effort was well below 1\,\%. The high reported measurement effort for smaller systems is due to our choice of the measurements per iterations, which vastly over-approximates. Since we are able handle systems with more configuration options with similar effort, we are confident that a smaller number of measurements per iteration would reduce the relative measurement effort required substantially.

@@

	For the experiments not terminating within limit (the rightmost bar in Figure~\ref{fig:results_rq1_frequency}), we conducted an additional analysis. We compared whether and how the parameter setting of Table~\ref{tab:synthetic_experiment_ranges} explains this non-termination. One setting stands out as the cause for not finishing the experiment within 30 iterations: the number of measurements per iterations. The lower the number of measurements ($n_\text{measurements}$), the more iterations our approach needs.
	\vspace{2mm}
	\greybox{
		\textbf{Summary ($\text{RQ}_1$)}: We are able to identify and pinpoint config\-uration-specific performance change points accurately and at scale. The number of measurements per iteration is the main factor influencing how fast our algorithm terminates.
	}

	\subsubsection{$\text{RQ}_2$ (Pre-Study)}
	For our three subject systems, we have manually identified a number of commits, for which performance chang\-ed substantially. We have found 7 change points for \textsc{lrzip}, 2 for \textsc{xz}, and 2 for \textsc{oggenc}. For \textsc{lrzip}, 5 of 7, and for \textsc{xz}, 2 of 2 performance changes affect multiple configurations. By contrast, 2 change points for \textsc{lrzip} affect all measured configurations. All of the measured configurations for \textsc{oggenc} show a shift in performance. That is, the identified change points for \textsc{oggenc} as well as the two for \textsc{lrzip} are likely not configuration-specific. 	
	To further understand possible relations of change points with configuration options, we searched in commit messages for clues. We discuss two notable findings.

@@

	Of the remaining 6 commits for \textsc{lrzip}, we identified one as a successor to a merge commit. Within the related pull request\footnote{https://github.com/ole-tange/lrzip/commit/1203a1853e892300f79da19f14aca11152b5b890}, we discovered that the configuration \textsc{lzma} was set to be ignored. This plausibly corresponds to an observed performance shift for configurations with this option enabled. However, this orphan configuration option results in an inconsistent configuration interface, which we discovered with our approach.

@@

	\section{Discussion}\label{sec:discussion}
	\paragraph{Efficiency} Among all experiments conducted with our synthetic setup, our approach yielded reasonable to high accuracy across most parameter settings. A subsequent analysis of parameter settings showed that our algorithm is not limited by the size of a problem space, but rather by the number of measurements per iteration. 
	Our algorithm found multiple change points across different parameter settings, including those with large numbers of commits and configuration options. 
	The vast majority of cases terminated in early iterations with high accuracy. That is, we were able to tackle a problem with exponential complexity (with respect to the number of configuration options from which we can draw new measurements) with a measurement budget that is linear in the number of iterations (we used only 30) and the measurements per iteration (between 100 and 500 measurements per iteration), respectively.
	Hence, we are confident that our approach scales and can efficiently approximate change points for large software systems. Notably, we are able to reproduce this finding in real-world settings when identifying the temporal location of change points. For our controlled experiment setup, we have not tuned the algorithms’ parameters to reach optimal efficiency or accuracy. However, even without tuning, we demonstrated the feasibility and efficiency of our approach. Project-specific fine tuning of several parameters is a possible extension to our approach and can further improve our results.

	\paragraph{Pinpointing Configuration Options}
	The synthetic setup has shown that our algorithm can conceptually handle both time and space. We were able to precisely identify commits for which performance changed substantially. By contrast, we were not able to fully reproduce this in a setting with real-world data when attempting to pinpoint change points to configuration options, which may be caused by an inconsistently documented configuration interface. Missing information in the real-world setup emphasizes the importance of a synthetic setup to evaluate approaches such as ours and explore possible limitations and necessary extensions. So, we have documented that, given a precisely identified commit, we can link the affected files to configuration options with little effort. In turn, given a set of code segments of a commit, matching this with the overall code base has been extensively studied before under the umbrella of feature location~\cite{hill_which_2013}. Inconsistencies in the documentation of configuration options is a well known and prevalent problem~\cite{rabkin_static_2011}. 
	Our approach might be a means to complement existing feature location techniques with information about configuration-dependent performance changes. 	
	In our latter experiment, instead of directly sampling new arbitrary configurations, we used a finite set of configurations as a proxy (cf.~Section~\ref{sec:rq2_operationalization}). This does not reflect the original layout of our algorithm, but lets us efficiently explore some limitations of this approach. We have learned that, for interpreting the results of our approach, domain knowledge is advantageous as we have identified incomplete and inconsistent configuration interfaces as possible limiting factors. We acknowledge that, while our approach performs well as an automated pipeline in a controlled environment, for real-world applications, a semi-automated setup with additional manual measurement prioritization and deeper knowledge of configuration constraints would improve the applicability of our approach. This is a promising further avenue of future work.

	\paragraph{Threats to Validity}
	Threats to \emph{internal validity} include measurement noise that may distort change point approximation. We mitigate this threat by repeating each measurement five times and reporting the mean. With respect to the mean performance, the reported variation was below 10\%. For RQ2, we considered only performance changes of, at least, 10\%. We conducted a pre-study, where we manually identified change points.
	Hence, we are confident that our raw data are robust against outliers. The setup in RQ2 relies on a limited set of previously sampled configurations instead of actively acquiring new ones. We mitigate this limitation by selecting a broad set of configurations in our pre-study with different sampling strategies. We sampled as many randomly selected configurations as there are pairs of configuration options, resulting in a reasonably high coverage of the configuration space.

	Regarding \emph{external validity}, we cannot claim that we have identified all limitations possibly arising in a practical setting. However, we selected popular software systems that are well optimized for performance and often make use of configuration options to tune performance.  Although we have shown possible limitations in a practical setting, our results, in conjunction with our exhaustive synthetic study, are reproducible, and we were able to test corner cases and assess scalability so that we believe that our results hold for many practical use cases.

>>>>

%%
%% This is file `sample-sigconf.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `sigconf')
%%
%% IMPORTANT NOTICE:
%%
%% For the copyright see the source file.
%%
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigconf.tex.
%%
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%%
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigconf, screen]{acmart}
\usepackage{microtype}
\usepackage[utf8]{inputenc}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{enumitem}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{mathtools} 
\usepackage{subcaption}
\usepackage{caption}
\usepackage{tikz}
\usepackage{mdframed}
\usepackage{wrapfig}
\usepackage{caption}
\usepackage{todonotes}
%\usepackage{hyperref}

@@

	\author{Stefan Mühlbauer}
	\affiliation{%
		\institution{Leipzig University}
		%\city{Doimukh}
		\country{Germany}}

	\author{Sven Apel}
	\affiliation{%
		\institution{Saarland University\\Saarland Informatics Campus}
		%\city{Doimukh}
		\country{Germany}}

	\author{Norbert Siegmund}
	\affiliation{%
		\institution{Leipzig University}
		%\city{Doimukh}
		\country{Germany}}
	%%
	%%
	%% By default, the full list of authors will be used in the page
	%% headers. Often, this list is too long, and will overlap
	%% other information printed in the page headers. This command allows
	%% the author to define a more concise list
	%% of authors' names for this purpose.
	%\renewcommand{\shortauthors}{Trovato and Tobin, et al.}

@@

	\begin{CCSXML}
		<ccs2012>
		<concept>
		<concept_id>10011007.10010940.10011003.10011002</concept_id>
		<concept_desc>Software and its engineering~Software performance</concept_desc>
		<concept_significance>500</concept_significance>
		</concept>
		<concept>
		<concept>
		<concept_id>10011007.10011074.10011111.10011113</concept_id>
		<concept_desc>Software and its engineering~Software evolution</concept_desc>
		<concept_significance>500</concept_significance>
		</concept>
		</ccs2012>
	\end{CCSXML}

	\ccsdesc[500]{Software and its engineering~Software performance}
	%\ccsdesc[500]{Computing methodologies~Active learning settings}
	\ccsdesc[500]{Software and its engineering~Software evolution}

@@

	%%
	%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
	%% Please copy and paste the code instead of the example below.h
	%%
	%% Keywords. The author(s) should pick words that accurately describe
	%% the work being presented. Separate the keywords with commas.
	\keywords{Software performance, software evolution, configurable software systems, machine learning, active learning}

@@

	\subsection{Performance-Influence Models}\label{sec:pimodels}
	Consider the following running example of a database management system (DBMS) with two selectable features: \textsc{Encryption} and \textsc{Compression}.
	Either feature adds execution time to the overall performance, but, if both features are selected, the execution time is smaller than the sum of the individual features' contribution to performance: Less data is encrypted if it is compressed beforehand.
	An \textit{interaction} in this setting is the combined effect of features \textsc{Encryption} and \textsc{compression}.
	We can assign each feature and interaction an \textit{influence} that it contributes to the overall performance, if selected. In our example, the individual influences of the two features are positive (increasing execution time), whilst the interaction's influence is negative (decreasing execution time).

@@

	We encode the threshold in a step function $\theta_\tau$:
	\begin{equation}\label{eq:theta}
	\begin{gathered}
	\theta_\tau(a, b) = \begin{cases}
	0,\quad \vert\,\pi(a) - \pi(b)\,\vert < \tau\\
	1,\quad \vert\,\pi(a) - \pi(b)\,\vert \geq \tau
	\end{cases}\quad a, b \in V, \tau \in \mathbb{R}
	\end{gathered}
	\end{equation}
	The function evaluates to $1$ if the difference between performance $\pi_a$ and $\pi_b$ of commits $a$ and $b$ exceeds the threshold $\tau$ and $0$ if not.

@@

	The threshold $t_\text{CPL}$ is the average change point likelihood over all commits plus a factor $N_\text{CPL}$ times its corresponding standard deviation over all commits. 
	The factor $N_\text{CPL}$ (default value: 3) allows us to filter peaks that do not stand out enough and might be false positives. The greater $N_\text{CPL}$ is, the stricter the filtering of peaks. That is, for each configuration in our learning set, we obtain a set of commits (configuration-dependent candidate commits) that represent possible change points.
	To reduce variation among the obtained commits, we cluster the commits using kernel density estimation (KDE). The purpose of KDE in our setting is to estimate the probability mass function of whether a commit is a change point. The local maxima of this KDE, subsequently, represent our set of candidate commits. Note that, if two distinct change points are almost coincident commit-wise, they can be mistakenly identified by our approach as one change point that, subsequently, can be associated with configuration options belonging to the two individual change points.

@@

	For each model (for a candidate commit), we consider a configuration option as associated with a commit, if its influence $c_i$ (i.e., the absolute value of its coefficient) is greater than a threshold:% $t_\text{influence}$:
	\begin{equation}
	\begin{gathered}
	t_\text{influence} = \frac{1}{n_\text{options}} + N_\text{influence} \cdot \underbrace{\sqrt{\frac{1}{n_\text{options}} \cdot \sum \Big(\frac{1}{n_\text{options}} - c_i\Big)^2}}_{\text{standard deviation of } c_i}
	\label{eq:t_influence}
	\end{gathered}
	\end{equation}
	We consider options as influential, if their estimated influence is greater than if each configuration option were equally influential. In addition, we add to this threshold the standard deviation of the influences of all configuration options. In theory, our regression model eliminates non-influential configuration options. If more than configuration option is influential (i.e., two or more options interact) their respective estimated influence should be roughly equal. To avoid false positives in he estimation of influential configuration options, we use the standard deviation and the parameter $N_\text{influence}$ (default value: 1) to filter only estimations with low variation among the estimated influences. If the intercept of a model (i.e., the coefficient not related to any regression model variable) exceeds this threshold, we consider this candidate commit a change point that is not configuration-dependent (i.e., it affects all configurations). The outcome is a list of \emph{associations}: pairs of candidate commits and (likely) influential configuration options. 
	% unklar, warum N_influence hier eingeführt wird

@@

	With this split experiment setup, we aim at answering the following two research questions:
	\begin{enumerate}[leftmargin=0.825cm]
		\item[\bf RQ$_1$:] Can we \emph{accurately} and \emph{efficiently} identify configuration-de\-pen\-dent performance change points? 
		\item[\bf RQ$_2$:] Can we \emph{practically} identify configuration-de\-pen\-dent change points \emph{in a real-world setting}? 
	\end{enumerate}
	% 
	\subsection{Controlled Experiment Setup}
	We break down the research question $\text{RQ}_1$ into three objectives  to study the influence of the size of configurable software systems, change point properties, and measurement effort.

@@

	\begin{table}
		\centering
		\captionof{table}{Parameter ranges for the synthetic experiment.}
		\label{tab:synthetic_experiment_ranges}
		\begin{tabular}{lll}
			\toprule
			\textbf{}                & \textbf{Parameter}        & \multicolumn{1}{l}{\textbf{Range}} \\
			\midrule
			\parbox[t]{20mm}{\multirow{4}{*}{\rotatebox[origin=c]{0}{\parbox{2cm}{\centering\footnotesize Synthesized Systems}}}} & $n_\text{options}$                & 8, 16, 32, 64                \\
			& $n_\text{commits}$                & 1000, 2500                \\
			& $n_\text{changepoints}$           & 1, 2, 5, 10                        \\
			& $p_\text{interaction}$ & 0.5, 0.7, 0.9                     \\
			\midrule
			\parbox[t]{20mm}{\multirow{2}{*}{\rotatebox[origin=c]{0}{\parbox{2cm}{\centering\footnotesize Initialization}}}}   & $N_\text{initial}$       & 2, 5, 10                           \\
			& $n_\text{measurements}$           & 100, 200, 500                      \\
			\bottomrule            
		\end{tabular}
	\end{table}

@@

	\begin{table}
		\centering
		\captionof{table}{Project characteristics of subject systems.}
		\label{tab:properties}
		\begin{tabular}{lrr}
			\toprule
			\textbf{Name} & \textbf{\#\,Options} & \textbf{\#\,Commits}\\
			\midrule
			\textsc{xz} & 16 & 1193\\
			\textsc{lrzip} & 9 & 743 \\
			\textsc{oggenc} & 12 & 935\\
			\bottomrule
		\end{tabular}
	\end{table}

@@

	\subsection{Results}	
	\subsubsection{$\text{RQ}_1$ (Controlled Experiment).}
	We illustrate the results of our first set of experiments in Figures~\ref{fig:results_rq1_accuracy_terminated}, \ref{fig:results_rq1_frequency}, and \ref{fig:results_rq1_degree}. The vast majority of experiments terminated within the limit of 30 iterations, as shown in Figure~\ref{fig:results_rq1_frequency}. A small portion of experiments, however, did not meet our termination criteria (more on that below). From all experiments that terminated in Figure~\ref{fig:results_rq1_accuracy_terminated}, we depict the F1 score after they have terminated, that is, after their last iteration\footnote{Due to space limitations, we report precision and recall on the paper's companion Web site.}. For most iterations, the mean F1 score falls around or over 0.7, with the first quartile  only slightly being as low as 0.5 (at iteration 29).  In the grand scheme of things, the vast majority of experiments terminated with reasonably high F1 score. 
	\begin{figure}
		\begin{subfigure}{\linewidth}
			\centering
			\includegraphics[width=\linewidth]{images/accuracy_after_termination.pdf}
			\vspace{-5mm}
			\caption{F1 score of experiments terminating at different iterations.}	
			\label{fig:results_rq1_accuracy_terminated}
		\end{subfigure}
		\begin{subfigure}{\linewidth}
			\centering
			\vspace{1mm}
			\includegraphics[width=\linewidth]{images/iterations_for_termination.pdf}
			\vspace{-5mm}
			\caption{Frequency of experiments terminating at different iterations.}	
			\label{fig:results_rq1_frequency}
		\end{subfigure}
		\begin{subfigure}{\linewidth}
			\vspace{4mm}
			\centering
			\includegraphics[width=\linewidth]{images/max_degree_after_termination}
			\caption{Influence of the number of change points and interaction degree on accuracy.}	
			\label{fig:results_rq1_degree}
		\end{subfigure}
		\caption{Result for the synthetic experiment.}
		\label{fig:results_rq1:results_rq1_degree}
	\end{figure} 	
	Regarding measurement effort, for small systems with 8 configuration options, we required up to 50\,\% of all possible measurements. For greater systems, the required measurement effort was well below 1\,\%. The high reported measurement effort for smaller systems is due to our choice of the measurements per iterations, which vastly over-approximates. Since we are able handle systems with more configuration options with similar effort, we are confident that a smaller number of measurements per iteration would reduce the relative measurement effort required substantially.

@@

	For the experiments not terminating within limit (the rightmost bar in Figure~\ref{fig:results_rq1_frequency}), we conducted an additional analysis. We compared whether and how the parameter setting of Table~\ref{tab:synthetic_experiment_ranges} explains this non-termination. One setting stands out as the cause for not finishing the experiment within 30 iterations: the number of measurements per iterations. The lower the number of measurements ($n_\text{measurements}$), the more iterations our approach needs.
	\vspace{2mm}
	\greybox{
		\textbf{Summary ($\text{RQ}_1$)}: We are able to identify and pinpoint config\-uration-dependent performance change points accurately and at scale. The number of measurements per iteration is the main factor influencing how fast our algorithm terminates.
	}

	\subsubsection{$\text{RQ}_2$ (Pre-Study)}
	For our three subject systems, we have manually identified a number of commits, for which performance chang\-ed substantially. We have found 7 change points for \textsc{lrzip}, 2 for \textsc{xz}, and 2 for \textsc{oggenc}. For \textsc{lrzip}, 5 of 7, and for \textsc{xz}, 2 of 2 performance changes affect multiple configurations. By contrast, 2 change points for \textsc{lrzip} affect all measured configurations. All of the measured configurations for \textsc{oggenc} show a shift in performance. That is, the identified change points for \textsc{oggenc} as well as the two for \textsc{lrzip} are likely not configuration-dependent. 	
	To further understand possible relations of change points with configuration options, we searched in commit messages for clues. We discuss two notable findings.

@@

	Of the remaining 6 commits for \textsc{lrzip}, we identified one as a successor to a merge commit. Within the related pull request\footnote{\url{https://github.com/ole-tange/lrzip/commit/1203a1853e892300f79da19f14aca11152b5b890}}, we discovered that the configuration \textsc{lzma} was set to be ignored. This plausibly corresponds to an observed performance shift for configurations with this option enabled. However, this orphan configuration option results in an inconsistent configuration interface, which we discovered with our approach.

@@

	\section{Discussion}\label{sec:discussion}
	\paragraph{Efficiency} Among all experiments conducted with our synthetic setup, our approach yielded reasonable to high accuracy across most parameter settings. A subsequent analysis of parameter settings showed that our algorithm is not limited by the size of a problem space, but rather by the number of measurements per iteration. 
	Our algorithm found multiple change points across different parameter settings, including those with large numbers of commits and configuration options. 
	The vast majority of cases terminated in early iterations with high accuracy. That is, we were able to tackle a problem with exponential complexity (with respect to the number of configuration options from which we can draw new measurements) with a measurement budget that is linear in the number of iterations (we used only 30) and the measurements per iteration (between 100 and 500 measurements per iteration), respectively.
	Hence, we are confident that our approach scales and can efficiently approximate change points for large software systems. Notably, we are able to reproduce this finding in real-world settings when identifying the temporal location of change points. For our controlled experiment setup, we have not tuned the algorithms’ parameters to reach optimal efficiency or accuracy. However, even without tuning, we demonstrated the feasibility and efficiency of our approach. Project-specific fine tuning of parameters may further improve the performance of our approach.

	\paragraph{Pinpointing Configuration Options}
	The synthetic setup has shown that our algorithm can conceptually handle both time and space. We were able to precisely identify commits for which performance changed substantially. By contrast, we were not able to fully reproduce this in a setting with real-world data when attempting to pinpoint change points to configuration options, which we attribute to missing information on and inconsistencies in the configuration interface.
	Missing information in the real-world setup emphasizes the importance of a synthetic setup to evaluate approaches such as ours and to explore possible limitations and necessary extensions. So, we have documented that, given a precisely identified commit, we can link the affected files to configuration options with little effort. In turn, given a set of code segments of a commit, matching this with the overall code base has been extensively studied before under the umbrella of feature location~\cite{hill_which_2013}. Inconsistencies in the documentation of configuration options is a well known and prevalent problem~\cite{rabkin_static_2011}. 
	Our approach complements existing feature location techniques with information on configuration-dependent performance changes. 	
	In our latter experiment, instead of directly sampling new arbitrary configurations, we used a finite set of configurations as a proxy (cf.~Section~\ref{sec:rq2_operationalization}). This does not reflect the original layout of our algorithm, but lets us efficiently explore some limitations of the approach. We have learned that, for interpreting the results of our approach, domain knowledge is advantageous, especially, as we have identified incomplete and inconsistent configuration interfaces as possible limiting factors. We acknowledge that, while our approach performs well as an automated pipeline in a controlled environment, for real-world applications, a semi-automated setup with additional manual measurement prioritization and deeper knowledge of configuration constraints would improve the applicability of our approach.

	\paragraph{Threats to Validity}
	Threats to \emph{internal validity} include measurement noise, which may distort change point approximation. We mitigate this threat by repeating each measurement five times and reporting the mean. With respect to the mean performance, the reported variation was below 10\%. For RQ$_2$, we considered only performance changes of, at least, 10\%. 
	%Furthermore, we conducted a pre-study, where we manually identified change points.
	Hence, we are confident that our raw data are robust against outliers. The setup in RQ$_2$ relies on a limited set of previously sampled configurations instead of actively acquiring new ones. We mitigate this limitation by selecting a broad set of configurations in our pre-study with different sampling strategies. We sampled as many randomly selected configurations as there are pairs of configuration options, resulting in a reasonably high coverage of the configuration space.

	Regarding \emph{external validity}, we cannot claim that we have identified all limitations possibly arising in a practical setting. However, we selected popular software systems that are well optimized for performance and often make use of configuration options to tune performance.  Although we have encountered possible limitations in a practical setting, our results, in conjunction with our exhaustive synthetic study, are reproducible, and we were able to test corner cases and assess scalability so that we believe that our results hold for many practical use cases.
